---
title: "Drawing chromosome idiograms with plot_ideogram()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Drawing chromosome idiograms with plot_ideogram()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 9, fig.align = "center"
)
```

```{r setup}
library(bioRtools)
```

`bioRtools` provides three ggplot2-based idiogram functions, reimplemented from
the [RIdeogram](https://cran.r-project.org/package=RIdeogram) package (Hao et
al. 2020, *PeerJ Comput. Sci.* 6:e251). Unlike RIdeogram — which writes SVG
files via hand-concatenated strings — these functions return **ggplot objects**,
so the result is themeable, composable with `patchwork`, and exportable to any
format via `ggsave()`.

This vignette walks through the single-species, dual-species, and three-species
(ternary) modes. Test data are bundled with the package.

## Single-species idiogram: `plot_ideogram()`

A karyotype is a data frame of `Chr, Start, End`, optionally with `CE_start,
CE_end` for the centromere (drawn as a constriction).

```{r karyotype-only, fig.height = 7}
data(df.ideo.karyotype)
head(df.ideo.karyotype)

plot_ideogram(df.ideo.karyotype)
```

### Overlaid heatmap

Pass an `overlaid` data frame (`Chr, Start, End, Value`) to fill each chromosome
body with a colour-mapped heatmap — e.g. gene density.

```{r overlaid}
data(df.ideo.gene_density)
plot_ideogram(df.ideo.karyotype, overlaid = df.ideo.gene_density)
```

### Annotation label tracks

The `label` argument adds a track to the right of the chromosomes. Four
`label_type`s are supported:

- **`"marker"`** — `Type, Shape, Chr, Start, End, color`; shapes (`triangle` /
  `box` / `circle`) with leader lines, vertically repelled to avoid overlap.
- **`"heatmap"`** — a second heatmap track (`Chr, Start, End, Value`), drawn
  with its own colour scale.
- **`"line"`** / **`"polygon"`** — line or filled-area tracks along the
  chromosomes.

```{r marker}
data(df.ideo.rna_marker)
plot_ideogram(df.ideo.karyotype,
              overlaid = df.ideo.gene_density,
              label = df.ideo.rna_marker,
              label_type = "marker")
```

Two heatmaps at once (overlaid + a heatmap label track) use independent colour
scales:

```{r two-heatmap}
data(df.ideo.ltr_density)
plot_ideogram(df.ideo.karyotype,
              overlaid = df.ideo.gene_density,
              label = df.ideo.ltr_density,
              label_type = "heatmap")
```

## Dual-species synteny: `plot_ideogram_synteny()`

Two species' chromosomes are drawn as horizontal rows with quadratic-Bezier
ribbons connecting syntenic blocks. The karyotype needs columns
`Chr, Start, End, fill, species, size, color` (two species); the synteny links
need `Species_1, Start_1, End_1, Species_2, Start_2, End_2, fill`.

```{r synteny, fig.width = 8, fig.height = 5}
data(df.ideo.synteny_karyotype)
data(df.ideo.synteny)
plot_ideogram_synteny(df.ideo.synteny_karyotype, df.ideo.synteny)
```

## Three-species (ternary) synteny: `plot_ideogram_ternary()`

Three species are arranged in a triangle — species 1 horizontal along the
bottom, species 2 rotated -60° (upper-left), species 3 rotated +60°
(upper-right) — with ribbons for each of the three pairwise comparisons. Set a
synteny row's `fill` to `"gradient"` to blend the two species' colours along
that ribbon.

```{r ternary, fig.width = 8, fig.height = 8}
data(df.ideo.ternary_karyotype)
data(df.ideo.ternary)
plot_ideogram_ternary(df.ideo.ternary_karyotype, df.ideo.ternary)
```

## Notes on the port

- Chromosome geometry matches the original RIdeogram formulas to full precision
  (validated against the original SVG coordinates). Ternary ribbons use a clean
  self-contained geometry, so their curvature is close but not pixel-identical
  to the original's hand-tuned curves.
- All three functions return ggplot objects — customize with `+ theme()`,
  combine with `patchwork`, or save with `ggsave("file.png", p, ...)`.
- Test datasets (`df.ideo.*`) are repackaged from RIdeogram under its Artistic-2.0
  license.
